---
title: "HW03"
author: "Donna Dalrymple"
date: "February 20, 2017"
output: 
  html_document:
      toc: true
      code_download: true
      theme: simplex
      toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE
  )
```

```{r libraries_we_need}
    library(knitr)
    library(readxl)
    library(dtables)
    library(ggplot2)
    library(tidyverse)
 ```
 
## Exploring Pipeline Incidents

Using data from <http://www.phmsa.dot.gov/pipeline/library/data-stats/flagged-data-files>
 
 ```{r get_data}
 URL <- "http://www.phmsa.dot.gov/staticfiles/PHMSA/DownloadableFiles/Pipeline/PHMSA_Pipeline_Safety_Flagged_Incidents.zip"
 fil <- "data/pipelines.zip"
 if (!file.exists(fil)) {
   download.file(URL, fil)
   unzip(fil, exdir="data")
 }
 data_file <- "data/hl2010toPresent.xlsx"
```
 
 Read in the data:
 
 ```{r load_data}
 pipelines <- read_excel(data_file, "hl2010toPresent", col_names=TRUE)
 ```
 
 ```{r eda, eval=FALSE, echo=FALSE}
 # take a look at the strucure
 glimpse(pipelines)
 ```
 
### Significant incidents

We want to identify if record meets the significant criteria or not:

- fatality
- injury
- fire
- explosion
- total property damage $50K or more in 1984 dollars
- non-HVL loss >= 50bbls
- HVL loss >= 5bbls

We'll do this by year.

```{r signigicant_incidents}
count(pipelines, IYEAR, SIGNIFICANT) %>% 
  arrange(IYEAR) %>% 
  spread(SIGNIFICANT, n) %>% 
  rename(`Incident year`=IYEAR,
         `Not Significant`=NO, 
         `Significant`=YES) -> sig_ind_by_year
```


`r kable(sig_ind_by_year)`

```{r sig_ind_plot}
gather(sig_ind_by_year, is_sig, ct, -`Incident year`) %>% 
  ggplot(aes(`Incident year`, ct, group=is_sig)) +
  geom_line(aes(color=is_sig)) +
  labs(title="Significant pipeline incidents by year (2010-2016)") +
  theme_bw() +
  theme(legend.position="bottom")
```

```{r sig_ind_plot_diff, fig.height=2}
mutate(sig_ind_by_year, diff=`Not Significant`- `Significant`) %>%
  ggplot(aes(`Incident year`, diff)) +
  geom_segment(aes(xend=`Incident year`, yend=0)) +
  labs(title="[Difference] Significant pipeline incidents by year (2010-2016)") +
  theme_bw() +
  theme(panel.grid.major.x=element_blank()) +
  theme(panel.grid.minor.x=element_blank())
```

### How about Serious Incident?
  These was either a fatality or an injury

```{r serious}
count(pipelines, IYEAR, SERIOUS) %>% 
  spread(SERIOUS, n, fill = 0) %>% 
  kable()
```

### How about Shutdowns?
Did the incident result in a shutdown?
```{r shutdown}
count(pipelines, IYEAR, SHUTDOWN_DUE_ACCIDENT_IND, sort=TRUE) %>%   filter(!is.na(SHUTDOWN_DUE_ACCIDENT_IND)) %>% 
  spread(SHUTDOWN_DUE_ACCIDENT_IND, n, fill = 0) %>% 
  kable()
```

### How about Spills?
Large and small spills by year

```{r spills}
count(pipelines, IYEAR, SPILL_TYPE_CATEGORY, sort=TRUE) %>% 
  spread(SPILL_TYPE_CATEGORY, n, fill = TRUE) %>% 
  kable()
```

### How about Fatalities and Injuries?
Fatalities and Injuries by year and type

```{r progress}
filter(pipelines,FATALITY_IND == "YES"| INJURY_IND == "YES") %>%
  select(matches("INJURIES|IYEAR|FATAL"),-FATALITY_IND, -INJURY_IND) %>%
  gather(category, value, - IYEAR)%>%
  rename(year = IYEAR)%>%
  count(year,category, wt = value)-> pipelinetrans

separate(pipelinetrans, category, into = c("num","type","fat_or_inj"), sep = "_", fill ="left") %>%
  select(-num)%>%
  filter(!is.na(type)) -> fatinj

fatinj %>%
  group_by(year, type)%>%
  mutate(tot = sum(n))%>%
  group_by(year, type, fat_or_inj)%>%
  mutate(pct = (n/tot))%>%
  arrange(year, type)%>%
  kable()
```

Sorted Injuries 
by quanity and type
```{r injuries}
fatinj %>%
  group_by(type, year)%>%
  filter(fat_or_inj == "INJURIES")%>%
  arrange(fat_or_inj, desc(n))%>%
kable()

```


Sorted Fatalities
by quanity and type
```{r fatalities}
fatinj %>%
  group_by(type, year)%>%
  filter(fat_or_inj == "FATALITIES")%>%
  arrange(type, desc(n))%>%
kable()

```


### How about On or Offshore?
Onshore vs offshore incidents by year

```{r locale}
count(pipelines, IYEAR, ON_OFF_SHORE, sort=TRUE) %>% 
  spread(ON_OFF_SHORE, n, fill = 0) %>% 
  mutate(tot = OFFSHORE + ONSHORE, off_pct = OFFSHORE/tot, on_pct = ONSHORE/tot  )%>%
  rename(year = IYEAR)%>%
  kable()
```

The table above shows pipeline incidents are more likely onshore.


## Exploring Pipe or Weld Failures

Provide a breakdown of incidents by type of pipe or weld failure 

```{r pipe_or_weld_failures}

pipelines %>%
  select(starts_with("PWF")) -> pwf_extract

cols <- colnames(pwf_extract)

pipelines[,c("REPORT_NUMBER",cols[-c(1:4,43)])]%>%
gather(indicator, indicator_value, ends_with("IND"), -REPORT_NUMBER)%>%
gather(year, year_value, ends_with("YEAR"), -REPORT_NUMBER)%>%
  select(report_num=REPORT_NUMBER, year=year_value, indicator, indicator_value) %>%
  filter(!is.na(indicator_value))%>%
  filter(!is.na(year)) %>%
  select(-indicator_value,pwf_indicator=indicator)%>%
  mutate(pwf_indicator=tolower(trimws(gsub("(^PWF_|_|IND)"," ",pwf_indicator)))) %>%
  distinct() -> pwf_df  
  

count(pwf_df, report_num) 


count(pwf_df, year, report_num) %>%
  ggplot(aes(n)) + geom_histogram()

count(pwf_df, pwf_indicator, sort=TRUE) %>%
  kable()
```

## Incidents by Year and Failure Type

```{r year and combined type}
pipelines %>% 
  group_by(IYEAR) %>%
 #filter(ITEM_INVOLVED %in% c("PIPE", "WELD, INCLUDING HEAT-AFFECTED ZONE")) %>%
  filter(!is.na(FAILURE_TYPE))%>%
  count(PIPE_TYPE, WELD_SUBTYPE, FAILURE_TYPE,IYEAR) %>%
  ungroup() %>%
  arrange(PIPE_TYPE, FAILURE_TYPE) %>%
  kable()

```
